Modelling the Lymphatic Metastatic Progression Pathways of OPSCC \\ from Multi-Institutional Datasets

### 2024-05-23 src/tex/ms.tex
[showyourwork] overleaf sync

We also need to introduce a separate collection of \glspl{rv} that describe the diagnostic observation of a patient's involvement. In analogy to the hidden true state $\mathbf{X}[t]$ at time $t$, we write this diagnosis as
%
\begin{equation}
    \mathbf{Z} = \left( Z_v \right) \qquad v \in \left\{ 1,2, \ldots, V \right\}
\end{equation}
%
We do not need to differentiate between different times $t$ here, since a patient is ever only diagnosed once, after which treatment usually starts timely.
Diagnosis and true state of a patient are formally connected via the sensitivity $s_N$ and specificity $s_P$ of the used diagnostic modality. In clinical practice, these modalities are \gls{ct}, \gls{mri}, or \gls{pet} scan, but it may also include information from biopsies after a \gls{fna} or other techniques to detect lymphatic metastases. For each \gls{lnl} $v$ the conditional probability table of $P\left( Z_v \mid X_v \right)$ looks like this:

@@

We define the time $t=0$ to be the moment a patient's tumor formed, and hence $X_v[t=0]=0 \,\,\, \forall v$. However, using this definition, we cannot know how many time-steps have passed until $t_\text{D}$, when the patient was diagnosed with cancer. We can only make the assumption that a patients with an earlier T-category tumor was \emph{probably} diagnosed after fewer time-steps than a patient with a advanced T-category tumor.

We can use this assumption by marginalizing over the diagnose times $t_\text{D}$ of patients in different T-categories using different prior distributions over the diagnose time. E.g., $P \left( t=t_\text{D} \mid \text{early} \right)$ for early T-category patients (T1 \& T2) and $P \left( t=t_\text{D} \mid \text{late} \right)$ for advanced T-category patients (T3 \& T4). Throughout this work we will use binomial distributions for these probability mass functions. They depend only on a single parameter, which when multiplied with the number of considered time-steps, conveniently also represents the distribution's mean.
%
\begin{equation} \label{eq:time_dist}
    P \left( t = t_\text{D} \mid \text{T}x \right) = \operatorname{\mathfrak{B}}(t_\text{max},p_{\text{T}x})
\end{equation}
%
Here, the parameter $p_{\text{T}x}$ can be interpreted as the probability that the patient with a tumor of T-category $x$ will be diagnosed at time-step $t+1$ given they are in time-step $t$. We will use as the latest time-step $t_\text{max} = 10$, which will therefore give us a distribution over the diagnosis time that has its mean at $\mathbb{E}[t_\text{D}] = 10 \cdot p_{\text{T}x}$.

@@

\begin{figure*}
    \script{bg_core_prevs.py}
    \begin{centering}
        \includegraphics[width=\textwidth]{figures/bg_core_prevs.png}
        \caption{Prevalence of involvement as predicted by the base graph model for different scenarios involving the most commonly metastatic \glspl{lnl} II, III, and IV (shaded histograms). The model's predictions are compared to Beta posteriors over the prevalence based on the frequency of the same scenarios in the data and a uniform prior (solid lines). The top panels of each of the three subfigures shows some selected scenarios with early T-category tumors and the bottom panel the same scenarios for advanced T-category. These figures show that for the most important \glspl{lnl} II, III, and IV the base graph model already fits the data well. \label{fig:bg_prevalences}}
    \end{centering}
\end{figure*}

@@

{\it Contralateral Spread:} So far, our model concerns itself with the ipsilateral lymphatic spread. However, our data suggests that at least contralateral \gls{lnl} II is frequently involved in \gls{opscc} patients. It is also of interest clinically to be able to decide when the contralateral side may be spared entirely, as this could meaningfully reduce side effects such as xerostomia and dysphagia [citation needed]. Therefore extending the model to include some or all contralateral \glspl{lnl} would represent an important addition on its way to a clinical validation and ultimately implementation.

>>>>

We also need to introduce a separate collection of \glspl{rv} that describe the diagnostic observation of a patient's involvement. In analogy to the hidden true state $\mathbf{X}[t]$ at time $t$, we write this diagnosis as
%
\begin{equation}
    \mathbf{Z} = \left( Z_v \right) \qquad v \in \left\{ 1,2, \ldots, V \right\}
\end{equation}
%
We do not need to differentiate between different times $t$ here, since a patient is ever only diagnosed once, after which treatment usually starts timely.
Diagnosis and true state of a patient are formally connected via the sensitivity $s_N$ and specificity $s_P$ of the used diagnostic modality. In clinical practice, these modalities are \gls{ct}, \gls{mri}, or \gls{pet} scan, but it may also include information from biopsies after a \gls{fna} or other techniques to detect lymphatic metastases. For each \gls{lnl} $v$ the conditional probability table of $P\left( Z_v \mid X_v \right)$ is given by:

@@

We define the time $t=0$ to be the moment a patient's tumor formed, and hence $X_v[t=0]=0 \,\,\, \forall v$. However, using this definition, we cannot know how many time-steps have passed until $t_\text{D}$, when the patient was diagnosed with cancer. We can only make the assumption that a patient with an earlier T-category tumor was \emph{probably} diagnosed after fewer time-steps than a patient with a advanced T-category tumor.

We can use this assumption by marginalizing over the diagnose times $t_\text{D}$ of patients in different T-categories using different prior distributions over the diagnose time. E.g., $P \left( t=t_\text{D} \mid \text{early} \right)$ for early T-category patients (T1 \& T2) and $P \left( t=t_\text{D} \mid \text{late} \right)$ for advanced T-category patients (T3 \& T4). Throughout this work we will use binomial distributions for these probability mass functions.
%
\begin{equation} \label{eq:time_dist}
    P \left( t = t_\text{D} \mid \text{T}x \right) = \operatorname{\mathfrak{B}}(t_\text{max},p_{\text{T}x})
\end{equation}
%
Here, the parameter $p_{\text{T}x}$ can be interpreted as the probability that the patient with a tumor of T-category $x$ will be diagnosed at time-step $t+1$ given they are in time-step $t$. We will use as the latest time-step $t_\text{max} = 10$. Binomial distributions depend only on a single parameter, which when multiplied with the number of considered time-steps, conveniently also represents the distribution's mean.\footnote{In our previous publication on the lymphatic progression model, we have shown that the shape of the time-prior and number of time-steps have no impact as long as only one T-category is considered \cite{ludwig_hidden_2021}.} $t_\text{max}=10$ will therefore give us a distribution over the diagnosis time that has its mean at $\mathbb{E}[t_\text{D}] = 10 \cdot p_{\text{T}x}$.

@@

\begin{figure*}
    \script{bg_core_prevs.py}
    \begin{centering}
        \includegraphics[width=\textwidth]{figures/bg_core_prevs.png}
        \caption{Prevalence of involvement as predicted by the base graph model for different scenarios involving the most commonly metastatic \glspl{lnl} II, III, and IV (shaded histograms). The model's predictions are compared to Beta posteriors over the prevalence based on the frequency of the same scenarios in the data assuming a uniform prior (solid lines). E.g., out 423 early T-category patients, 100 (23.6\%) presented with involvement of the \glspl{lnl} II and III, as shown by the blue line in the uppermost panel. The top panels of each of the three subfigures shows some selected scenarios with early T-category tumors and the bottom panel the same scenarios for advanced T-category. These figures show that for the most important \glspl{lnl} II, III, and IV the base graph model already fits the data well. \label{fig:bg_prevalences}}
    \end{centering}
\end{figure*}

@@

{\it Contralateral Spread:} The work presented in this paper considers only ipsilateral lymphatic spread. To guide the elective CTV-N definition for the contralateral neck, the model must be extended to the contralateral side. Contralateral lymph node involvement is strongly dependent on whether the primary tumor extends over the midsaggital plane, but also on T-category and the extent of ipsilateral involvement \cite{ludwig_detailed_2022}. A comprehensive model accounting for these risk factors on contralateral spread is the subject of a follow-up publication.
